### 2024-03-07 tex/lrsub.tex
[ correction ] first pass over chapter 5

I can now show that, using the notion of \emph{environment} derived in
\cref{sec:lrkits}, we can replicate the Agda proofs from
\cref{sec:syntactic-kits} in the usage-aware setting of $\name$.
From \cref{sec:lenv}, we know that environments are preserved under all
syntax-forming operations: zero, addition, scaling, and binding.
What is left is to show how these properties are deployed, and also how to
go on and prove the admissibility of simultaneous renaming, simultaneous
substitution, and then single substitution.

@@

We start with a slightly modified definition of \AgdaRecord{Kit}.
We saw in \cref{thm:lr-bind} that in the usage-annotated context, we restrict
weakening of $\V$-values to just $\gr0$-use variables.
Meanwhile, the function $\mathrm{vr}$, also seen in \cref{thm:lr-bind}, maps
usage-checked variables to $\V$-values, and the function $\mathrm{tm}$, used
to coerce $V$-values yielded by the environment into terms, stays the same.
I state weakening in a slightly different way than previously, so as to help
unification against a known result type (avoiding the problem described by
\citet{McBride12} as \emph{green slime}).
The type \AgdaFunction{Weakening}\AgdaSpace{}\AgdaBound{$\V$} can be read as
saying that, for any context $\grP\gamma$ of shape $s + t$, if the right of
$\grP$ is below $\gr0$, then a value in the left part of $\grP\gamma$ weakens
to a value in the whole of $\grP\gamma$.

@@

Given \AgdaFunction{bindEnv} (\cref{thm:lr-bind}), \AgdaFunction{env-+}
(\cref{thm:lr-env-add}), and \AgdaFunction{env-*} (\cref{thm:lr-env-scale}),
we can reproduce the syntactic traversal \AgdaFunction{trav}.
With all these lemmas in place, writing \AgdaFunction{trav}
becomes routine.
When processing a rule, we work our way up through the
premise connectives, applying \AgdaFunction{env-*} wherever we see a
\AgdaFunction{$\cdot^c$}, \AgdaFunction{env-+} wherever we see a
\AgdaFunction{$*^c$}, and \AgdaFunction{bindEnv} wherever we see a
\AgdaFunction{Bind}.
We then use whatever environments (with names beginning with
\AgdaBound{$\rho$}) and whatever usage vector splitting facts (with names
beginning with \AgdaBound{sp}) come out of this process to recursively
traverse the subterms and recombine the results.

>>>>

I now show that, using the notion of \emph{environment} derived in
\cref{sec:lrkits}, we can replicate the Agda proofs from
\cref{sec:syntactic-kits} in the usage-aware setting of $\name$.
From \cref{sec:lenv}, we know that environments are preserved under all
syntax-forming operations: zero, addition, scaling, and binding.
What is left is to show how these properties are deployed, and also how to
go on and prove the admissibility of simultaneous renaming, simultaneous
substitution, and then single substitution.

@@

I start with the definition \AgdaFunction{Weakening}, which says what it means
for a family of values \AgdaBound{$\V$} to respect one step of weakening on the
right by $\gr0$-use variables.
I state weakening in a slightly different way to what appears in the statement
of \cref{thm:lr-bind}, so as to help
unification against a known result type (avoiding the problem described by
\citet{McBride12} as \emph{green slime}).
The type \AgdaFunction{Weakening}\AgdaSpace{}\AgdaBound{$\V$} can be read as
saying that, for any context $\grP\gamma$ of shape $s + t$, if the right of
$\grP$ is below $\gr0$, then a value in the left part of $\grP\gamma$ weakens
to a value in the whole of $\grP\gamma$.

\ExecuteMetaData[\LRKits]{Weakening}

Given this new definition of \AgdaFunction{Weakening}, the record
\AgdaRecord{Kit} remains largely unchanged relative to what we saw in
\cref{sec:syntactic-kits}.
We still want $\V$-values to respect weakening (as used in \cref{thm:lr-bind}),
and to support maps \AgdaField{vr} from variables (also used in
\cref{thm:lr-bind}) and \AgdaField{tm} into terms (used when traversal reaches a
variable, we get a value from the environment, and want to produce a term from
that value).
As well as \AgdaFunction{Weakening}, note that \AgdaRecord{\_$\sqni$\_} and, of
course, \AgdaDatatype{\_$\vdash$\_} have different definitions to the
corresponding intuitionistic notions, but they still represent morally the same
parts of the language and its metatheory.

@@

Given \AgdaFunction{bindEnv} (\cref{thm:lr-bind}), \AgdaFunction{env-+}
(\cref{thm:lr-env-add}), and \AgdaFunction{env-*} (\cref{thm:lr-env-scale}),
we can reproduce the syntactic traversal \AgdaFunction{trav}.
Similarly to the unchanged high-level definition of \AgdaRecord{Kit}, we are
aiming for an unchanged traversal principle expressed by the rule below.
When $\V$ has a \AgdaRecord{Kit} structure, and we have a $\V$-environment from
$\Gamma$ to $\Delta$, we can transform a term in $\Delta$ to a term in $\Gamma$
of the same type.

\[
  \begin{prooftree}
    \hypo{\text{\AgdaRecord{Kit}\AgdaSpace{}}\V}
    \hypo{\Gamma \env\V \Delta}
    \hypo{\Delta \vdash A}
    \infer3[Trav]{\Gamma \vdash A}
  \end{prooftree}
\]

With all the lemmas of \cref{sec:lenv} in place, writing \AgdaFunction{trav}
becomes routine.
When processing a rule, we work our way up through the
premise connectives, applying \AgdaFunction{env-*} wherever we see a
\AgdaFunction{$\cdot^c$}, \AgdaFunction{env-+} wherever we see a
\AgdaFunction{$*^c$}, and \AgdaFunction{bindEnv} wherever we see a
\AgdaFunction{Bind}.
We then use whatever environments (with names beginning with
\AgdaBound{$\rho$}) and whatever usage vector splitting facts (with names
beginning with \AgdaBound{sp}) come out of this process to recursively
traverse the subterms and recombine the results.

### 2024-03-07 tex/rensub.tex
[ correction ] first pass over chapter 5

In \cref{sec:semirings}, I defined my calculus of interest $\name$.
In this chapter, I develop the necessary syntactic metatheory for
specifying and implementing the substitution operation.
I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
significant changes to the underlying notion of \emph{environment} before doing
so.
I give and informally motivate these changes to environments in
\cref{sec:lrkits}, and prove some properties of the new definition in
\cref{sec:lenv}.
Finally, I apply these new environments to the syntax of $\name$ in
\cref{sec:lrsub} to derive renaming and substitution operators.

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Comparison with Petricek's substitution lemma}\label{sec:petricek}
\input{petricek.tex}
\section{Conclusion}\label{sec:ren-sub-lr-conc}

>>>>

In \cref{sec:semirings}, I defined my calculus of interest $\name$.
In this chapter, I develop the necessary syntactic metatheory for
specifying and implementing the simultaneous substitution operation.
My aim in this chapter is to replay the development in \cref{sec:kits}, making
minimal changes to lemmas and theorems so as to support semiring usage
annotations.
The big change I make in this chapter is to the definition of
\emph{environment}.
In terms of ideas, I believe that the definition of environment I give in
\cref{sec:lrkits} is the main contribution of this thesis.
Its primary novel feature is the use of a linear map --- in the sense found in
linear algebra --- to distribute usage annotations across the values in the
environment.

The resulting substitution operation bears some similarity to a multi-variable
substitution operation given by \citet{petricek-thesis}.
This is the only other presentation I have found in the literature of an
operation similar to simultaneous substitution for a linear or linear-like
calculus.
I discuss Petricek's substitution lemma in more detail in \cref{sec:petricek}.
In brief, relative to Petricek's work, I extract the environment from the
substitution lemma, tidy up the definition, and present usage-annotated
environments as a reusable tool and object of study in their own right.
I reuse the resulting definition of environment in the remainder of this thesis,
taking advantage of the properties and generality of definition established in
this chapter, particularly in \cref{sec:semantics,sec:example-semantics}.

The body of this chapter starts with a motivation for and a definition of
environment in the usage-annotated setting in \cref{sec:lrkits}.
I then give some key properties of and operations on these environments in
\cref{sec:lenv}.
I apply the previous definitions and properties to the problems of simultaneous
renaming and substitution for $\name$ in \cref{sec:lrsub}, which also connects
the previous sections' mathematical developments with Agda code.
I then compare the yielded substitution operation with the one given by Petricek
in \cref{sec:petricek}, before a concluding discussion in
\cref{sec:ren-sub-lr-conc}.

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Comparison with Petricek's substitution lemma}\label{sec:petricek}
\input{petricek.tex}

\section{Conclusion}\label{sec:ren-sub-lr-conc}
